BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-03-27 Paper/D5.5/D5.5.tex
Minor indent fix.
Small fix before sending to BSC.
Added Harald Servat corrections.

\bibliographystyle{plain}
\begin{document}
\devnum{5.5}
\title{BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications}
\version{0.3}
\deadline{18}
\level{RE}
\nature{Report}
\coordinator{CNRS}
\contributors{Thierry Deutsch (CEA), Luigi Genovese (CEA), Jean-François Méhaut (UJF), Kevin Pouget (CNRS), Brice Videau (CNRS)}
\reviewers{Olivier Aumage (INRIA), Frédéric Desprez (INRIA)}
\keywords{BOAST, Deliverable}

@@

\begin{changelog}
\change{0.1}{Initial version of the deliverable}
\change{0.2}{First version sent to reviewers}
\change{0.3}{With corrections from reviewers}
\end{changelog}

@@

\begin{executive}
This document describes BOAST, a metaprogramming framework to produce portable
and efficient computing kernels for HPC application. BOAST offers an embedded
domain specific language to describe the kernels and their possible
optimization. BOAST also supplies a complete runtime to compile, run, benchmark,
and check the validity of the generated kernels. BOAST is being used in two
flagship HPC applications BigDFT and SPECFEM3D, to improve performance
portability of those codes.
\end{executive}

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources. Nonetheless it is a very important aspect of the
Mont-Blanc project. Indeed, for the project, more than 10 applications were
selected to be ported and optimized for the prototype platform.

@@

Genericity of HPC codes is often limited. One of the reason is that producing
generic code in FORTRAN 90/95 is difficult as the language is not really fit
for it. Sometimes, adding genericity will degrade performance as optimization
opportunities that come from over-specification are lost.

@@

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in FORTRAN for historical and
performance reasons. Codes can be quite huge (several thousands lines of code)
with lots of functionalities. Nonetheless they are usually based on computing
kernels.  Computing kernels are resource intensive and well defined parts of a
program that usually work on precisely defined data. Those kernels are the time
consuming part of an HPC application. They are thus the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes some
of those developers only work a few monthes on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
FORTRAN may sometimes arrive later.

@@

What we want is computing kernels that are:
\begin{itemize}
\item Written in a \emph{portable} manner,
\item Written in a way that raises developer \emph{productivity},
\item Written to present good \emph{performance}.
\end{itemize}

@@

\paragraph{Vectorization}
The first proposed optimization is to compute 5 pixels instead of one using the
vectors available to the Mali architecture. To be efficient the vectors have to
be directly loaded from memory. Using vectors of 16 elements 15 useful components
are simultaneously loaded and can be used in computation. This optimization
yields speedups between 1.5 and 6 depending on the image size.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{BOAST_Workflow.pdf}
\caption{Structure and Workflow of the BOAST framework.}
\label{fig:boast_workflow}
\end{center}
\end{figure}

The ideas behind BOAST were first presented in~\cite{videau2013boast}, and are
to provide scientific application developers with a framework to develop and
test application computing kernels. The envisioned workflow and program
structure is presented in Figure~\ref{fig:boast_workflow}.  The user should
start from an application kernel (implemented or just designed) and implement it
in a dedicated language (step 1). The language provides enough flexibility for
the kernel to be meta-programmed with several orthogonal optimizations. Then the
developer selects one configuration of the kernel and a target language. Those
parameter define the output source code that will be generated by BOAST (step
2). The resulting code source is then built according to the user specified
compiler and options (step 3). If input data are available then the kernel can
be benchmarked and tested for non regression. Based on the results, other
optimizations can be selected (step 4) or new optimizations can be added to the
BOAST sources. The process can be repeated until a good candidate is found on
the target platform. The resulting kernel is then added to the program (step 5).

Several steps in the workflow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking... This automation can be
done by scripting by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution runtime.
\end{itemize}

@@

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite FORTRAN and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions must be composed in complex
Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

@@

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance commonly used operators have to be available
and behave as expected. It must also be possible to differentiate an action on
an abstraction in the context of an expression and in the context of the
management of this expression. $c = a + b$, an expression that affects the
results of $a + b$ to $c$  is not equal to $c \leftarrow a + b$, which saves the
expression $a + b$ to a variable $c$. This is why an embedded domain specific
language approach was selected~\cite{hudak1996building}. This allows for the
coexistence of two languages: the host language and the DSL. In our case the DSL
allows the description of the kernel ($c = a + b$) while the host language
provides the meta programming of the kernel ($c \leftarrow a + b$)%, cf Figure~\ref{fig:dsl}
. Operator overloading of the host language will provide the familiar syntax
programmers are accustomed to. The host language will also provide easy
interfacing with libraries we might need during the development of our framework.

It was also important to have our constructs like \emph{for} loops to have a
syntax approaching those commonly found in programming languages. For this we
needed a language which could seamlessly pass a block of code to a function.
Ruby~\cite{matsumoto2002ruby} is one such language. It has deep introspection
capabilities as well. This is the main reason why it was selected.

    \subsubsection{BOAST keywords}

@@

\begin{minipage}[b]{0.50\linewidth}
\centering
\lstset{style=customf,
        caption={FORTRAN output.},
        label={lst:BOAST_algebra_FORTRAN} }

@@

\begin{lstlisting}
int32_t i;
int64_t k;
double l[27];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
l[k + 1 - (-5)] = sin(i + 5);
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from FORTRAN. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It is used to combine variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator. Function calls
(FuncCall) are also abstracted and can be used in Expressions.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to ruby
variables of the corresponding name.

On Line~4 we declare those three variables. Variable \emph{i} is then affected
the value 5. On Line~6 the \emph{j} ruby variable is used to store the BOAST
expression \emph{$i+5$}. This variable will the be used transparently through
the rest of the program.

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
comma separated, similar to FORTRAN notation.

@@

\begin{minipage}[b]{0.47\linewidth}
\centering
\lstset{style=customf,
        caption={FORTRAN output.},
        label={lst:BOAST_control_FORTRAN} }

@@

The classical control structures are implemented. \emph{If}, \emph{For},
\emph{While}, \emph{Case} are abstractions in BOAST matching the behavior of
corresponding control structures in other languages. An exception is the For in
BOAST more closely matches the for in FORTRAN than the one in C.

Figure~\ref{fig:BOAST_control} shows some basic usages of the control structures.
The example shows a C macro or function that behaves similarly to the FORTRAN modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

\lstset{style=customf,
        caption={FORTRAN output.},
        label={lst:BOAST_procedure_FORTRAN} }

@@

The last control structure is Procedure. It is used to describe either procedure
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It is the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those arrays have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

The generated outputs in FORTRAN~\ref{lst:BOAST_procedure_FORTRAN} and
C~\ref{lst:BOAST_procedure_C} show the difference in meta-information that is
kept between both versions.

@@

\subsection{BOAST Runtime}

In the previous Section we presented BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby. But in
order to find the best version of a computing kernel, each version has to be
compiled, linked and then executed to assess its performance. This can be very
time consuming if this process cannot be automated. By enabling more versions to
be evaluated the automation will bring improved portability, better performance,
and in the end will improve the productivity of the developer.

In this section we will present the different aspects of BOAST runtime that
allow this automation. Those aspects are listed here:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance can vary between platforms. It is thus
important to be able to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages can have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

@@

Compilation of the generated kernels must also be very flexible. Indeed, HPC
applications may encounter platforms with very diverse compilation environments.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system exhibits similar behavior than common build systems. Compilers
and their compile/build options can be specified. This can be done at several
places. Here is the list by increasing order of precedence: BOAST configuration
file, environment variables, and at kernel build time.

@@

The next logical step is to benchmark the built kernel. BOAST offers a simple
way to run a kernel that was successfully built while staying inside of BOAST.
A built BOAST \emph{CKernel} exposes a run method that accepts arguments corresponding
to the BOAST \emph{Procedure} used to instantiate the kernel. Arrays must be
instances of \emph{NArray} which are numerical arrays that use C arrays
underneath.

Arrays which correspond to output parameters will be modified during the
execution of the kernel so results can be checked. This allows for easy non
regression testing. The run method also returns information (and result for
kernels that are functions as well as output scalars) about the run. For
instance the run time of the kernel is one such information, it is obtained
using the system wide realtime clock. Other probes can be inserted at compile
time if needed.

PAPI~\cite{mucci1999papi} is supported and different counters can be specified
during each execution of a kernel.

@@

The non regression methodology presented before is viable provided input data
can be generated at runtime. For instance in the case of the Laplace kernel,
generating an input image and the corresponding reference output image, using a
reference implementation, is easy. Unfortunately it is not always possible, some
applications have very complicated data patterns that are difficult to
synthesize without running the full application. Thus BOAST offers a way to load
binary data from the file system and use them as inputs of a kernel. Outputs can
also be checked against those binary data thus enabling (almost) data oblivious
non regression testing.

@@

Debugging applications running on GPU environments is well-recognized as a hard
and time-consuming activity. In complement with BOAST, we designed a trace-based
debugging tool that simplifies this porting operation. The tool relies on BOAST
support of multi-target code generation (Section~\ref{sec:multitarget}), used to
validate an application from one GPU-programming framework to another or between
different implementations using the same programming model. A case-study of the
port of a Cuda application to OpenCL is presented in
Section~\ref{subsec:specfem}.

@@

However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of API
parameters are handles to opaque types, so, in order to provide
meaningful information, \code{GPUTrace} stores information about these
objects (handle value, buffer creation size and attributes, kernel
name and prototype, \etc{}). This is a custom trace format.

Once two call traces have been generated by \code{GPUTrace}, it is
easy to compare them with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}

By default, \code{GPUTrace} only prints a unique identifier (the
creation index) for the memory buffers. A first flag can be activated
to append the first bits of the buffer to the trace, printed in the
right format, for visual inspection. Or another flag can be set to
dump the whole content of the buffer into a file, for a full
inspection.

@@

In this section we will present the benefits of using BOAST on the Laplace
motivating example as well as two scientific applications that use BOAST. The
first one, BigDFT~\cite{genovese2008daubechies}, uses boast in order to develop new
functionalities with performance portability in mind. The second one,
SPECFEM3D~\cite{komatitsch2011fluid}, uses BOAST to factorize OpenCL and CUDA
development, while having robust non regression tests.

@@

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimizations can have an impact on the
performance of the kernel on the Mali architecture. But what of other
architectures? And are there other optimizations that can impact the
performance?

@@

Here are the parameters we finally selected for our kernel optimization and
their possible values:
\begin{itemize}
\item \textit{x\_component\_number}: a positive integer
\item \textit{y\_component\_number}: a positive integer
\item \textit{vector\_length}: 1, 2, 4, 8 or 16
\item \textit{temporary\_size}: 2 or 4
\item \textit{synthesize\_loads}: \textit{true} or \textit{false}
\item \textit{vector\_recompute}: \textit{true} or \textit{false}
\end{itemize}
The last parameter is used when \textit{x\_component\_number} is not divisible by
\textit{vector\_length}. Two solutions are possible then, divide the remainder
of the division in vectors of smaller sizes (\textit{vector\_recompute} =
\textit{false}) or load more data and compute useless values in vectors of the
specified size. This last option is mimicking the behavior of ARM
implementation, although when working at the component level it may seem an
absurd thing to try.

@@

Table~\ref{tbl:ARM} are the best results obtained by ARM on different images
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically beats the hand
optimized version. As far as the optimization options are concerned the result
are a disappointment: the same kernel configuration is the best for all image
sizes. This kernel is using \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} has no meaning here as \textit{x\_component\_number}
== \textit{vector\_length}. Those results also shows that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in 1 cycle without
compromising the bandwidth. The results shown here are the best of 4 runs for
each configuration. As the test environment was well controlled kernel
durations were very stable.

@@

But what if we run our benchmark on other architectures? Table~\ref{tbl:other}
shows the results obtained when running our BOAST implementation on other
architectures. The chosen architectures were an Intel i7-2760QM CPU (Sandy
Bridge architecture) that supports OpenCL 1.2 and an NVIDIA gtx680 GPU that
support OpenCL 1.1. We can see that the performance ratio between the different
architectures is rather stable across image sizes.

The optimization parameters selected are not the same for those architectures.
Indeed, the Intel CPU favors kernels that have the parameters:
\textit{x\_component\_number} = 16,
\textit{vector\_length} = 8,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false. Once again
\textit{vector\_recompute} has no meaning. \textit{y\_component\_number} varies
from 4 to 2 when image size increase, thus decreasing task granularity as the
global work size increase. This result is unexpected and understanding why this
happen could be interesting. The NVIDIA GPU favors processing square tiles:
\textit{x\_component\_number} = 4, \textit{y\_component\_number} = 4,
\textit{vector\_length} = 4, \textit{temporary\_size} = 2 and
\textit{synthesize\_loads} = false. Once more \textit{vector\_recompute} has no
meaning.

>>>>

\bibliographystyle{plain}
\begin{document}
\devnum{5.5}
\title{BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications}
\version{0.5}
\deadline{18}
\level{RE}
\nature{Report}
\coordinator{CNRS}
\contributors{Thierry Deutsch (CEA), Luigi Genovese (CEA), Jean-François Méhaut (UJF), Kevin Pouget (CNRS), Brice Videau (CNRS)}
\reviewers{Olivier Aumage (INRIA), Frédéric Desprez (INRIA), Harald Servat (BSC)}
\keywords{BOAST, Deliverable}

@@

\begin{changelog}
\change{0.1}{Initial version of the deliverable}
\change{0.2}{First version sent to reviewers}
\change{0.3}{With corrections from reviewers}
\change{0.4}{Additional corrections from reviewers}
\end{changelog}

@@

\begin{executive}
This document describes BOAST, a metaprogramming framework to produce portable
and efficient computing kernels for HPC application. BOAST offers an embedded
domain specific language to describe the kernels and their possible
optimization. BOAST also supplies a complete run-time to compile, run, benchmark,
and check the validity of the generated kernels. BOAST is being used in two
flagship HPC applications BigDFT and SPECFEM3D, to improve performance
portability of those codes.
\end{executive}

@@

Porting and tuning HPC applications to new platforms is tedious and costly in
terms of human resources. Nonetheless, it is a very important aspect of the
Mont-Blanc project. Indeed, for the project, more than ten applications were
selected to be ported and optimized for the prototype platform.

@@

Genericity of HPC codes is often limited. One of the reason is that producing
generic code in Fortran 90/95 is difficult as the language does not really fit
for it. Sometimes, adding genericity degrades performance as optimization
opportunities that come from over-specification are lost.

@@

Scientific Computing Applications are usually developed by physicist, chemist or
meteorologist. Those codes are usually written in Fortran for historical and
performance reasons. Codes can be quite huge (several thousands lines of code [LOC])
with lots of functionalities. Nonetheless, they are usually based on computing
kernels.  Computing kernels are resource intensive and well-defined parts of a
program that usually work on precisely defined data. Those kernels represent the most
time-consuming part of an HPC application, and consequently, they are the prime target for
optimization.

Those applications are often developed by several individuals. Sometimes, some
of those developers only work a few months on the application. Maintaining
optimized code written by someone else is quite a challenge. Several languages
and programming paradigms can also be used in a project and thus the maintainer
must be knowledgeable in several areas of expertise. Portability problems can
also be caused by the availability of an optimizing compiler for a specific
language and architecture. C compilers are usually the first available while
Fortran may sometimes arrive later.

@@

What we want is computing kernels that are written:
\begin{itemize}
\item in a \emph{portable} manner,
\item in a way that raises developer \emph{productivity},
\item and, present good \emph{performance}.
\end{itemize}

@@

\paragraph{Vectorization}
The first proposed optimization involves computing five pixels instead of one using the
vectors available to the Mali architecture. The vectors have to reside in memory to
ensure an efficient execution. Using vectors of 16 elements, 15 useful components
are simultaneously loaded and can be used in computation. This optimization
yields speedups between 1.5 and 6 depending on the image size.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{BOAST_Workflow.pdf}
\caption{Structure and Work-flow of the BOAST framework.}
\label{fig:boast_workflow}
\end{center}
\end{figure}

BOAST provides scientific application developers with a framework to develop and
test application computing kernels~\cite{videau2013boast}.
Figure~\ref{fig:boast_workflow} illustrates the envisioned work-flow and program
structure.
The user starts from an application kernel (either designed or implemented), and
write it in a dedicated language (step 1).
The language provides enough flexibility for
the kernel to be meta-programmed with several orthogonal optimizations. Then the
developer selects a kernel configuration and a target language. Those
parameters define the output source code that will be generated by BOAST (step
2). The resulting code source is then built according to the user specified
compiler and options (step 3). If input data are available then the kernel can
be benchmarked and tested for non regression. Based on the results, other
optimizations can be selected (step 4) or new optimizations can be added to the
BOAST sources. The process can be repeated until a good candidate is found on
the target platform. The resulting kernel is then added to the program (step 5).

Several steps in the work-flow can be automated: optimization and compiler flags
exploration, non regression testing, benchmarking... This automation can be
scripted by the user or by interfacing other dedicated tools. In order
to achieve those goals three aspects should be considered:
\begin{itemize}
\item code description,
\item code generation,
\item and kernel execution run-time.
\end{itemize}

@@

BOAST kernel description language should be able to express all these
optimizations. This gives us a set of constraints to implement in the language:
\begin{itemize}
\item Arbitrary number of variables have to be created and manipulated (types,
attributes...).
\item Procedures have to be abstracted (reunite Fortran and C like languages,
attributes...).
\item Functions must be available.
\item Variables, constants and functions must be composed in complex
Expressions.
\item Basic control structures (for, while, if/else...) have to be abstracted.
\item Powerful array management (several dimensions, transformations,
indexing...).
\end{itemize}

@@

In order to manipulate those abstractions we want to have a syntax resembling
what programmers use. For instance, commonly used operators have to be available
and behave as expected. It must also be possible to differentiate an action on
an abstraction in the context of an expression and in the context of the
management of this expression. $c = a + b$, an expression that affects the
results of $a + b$ to $c$  is not equal to $c \leftarrow a + b$, which saves the
expression $a + b$ to a variable $c$. This is why an embedded domain specific
language approach was selected~\cite{hudak1996building}. This allows for the
coexistence of two languages: the host language and the Domain Specific Language (DSL).
In our case, DSL allows the description of the kernel ($c = a + b$) while the host language
provides the meta programming of the kernel ($c \leftarrow a + b$)%, cf Figure~\ref{fig:dsl}
. Operator overloading of the host language will provide the familiar syntax
programmers are accustomed to. The host language will also provide easy
interfacing with libraries we might need during the development of our framework.

It was also important to have our constructs like \emph{for} loops to have a
syntax approaching those commonly found in programming languages. To this end, we
needed a language which could seamlessly pass a block of code to a function.
Ruby~\cite{matsumoto2002ruby} is one such language. It has deep introspection
capabilities as well. This is the main reason why it was selected.

    \subsubsection{BOAST Keywords}

@@

\begin{minipage}[b]{0.50\linewidth}
\centering
\lstset{style=customf,
        caption={Fortran output.},
        label={lst:BOAST_algebra_FORTRAN} }

@@

\begin{lstlisting}
int32_t i;
int64_t k;
double l[27];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
l[k + 1 - (-5)] = sin(i + 5);
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from Fortran. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It combines variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the Ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator. Function calls
(FuncCall) are also abstracted and can be used in Expressions.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{pr} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
produces the Fortran (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. At first we define 2 Variables
\emph{i} and \emph{k} (note that \emph{k} is $64$ bit integer). The third
variable named \emph{l} is a one dimensional local array of length $27$, it
is indexed in the range $-5$ to $21$. All those Variables are affected to Ruby
variables of the corresponding name.

On Line~4 we declare those three variables. Variable \emph{i} is then affected
the value 5. On Line~6 the \emph{j} Ruby variable is used to store the BOAST
expression \emph{$i+5$}. This variable will the be used transparently through
the rest of the program.

On Line~8 we use the \emph{k} variable to index into the array \emph{l} using
the bracket operator. Would the array be multidimensional the index would be
comma separated, similar to Fortran notation.

@@

\begin{minipage}[b]{0.47\linewidth}
\centering
\lstset{style=customf,
        caption={Fortran output.},
        label={lst:BOAST_control_FORTRAN} }

@@

The classical control structures are implemented. \emph{If}, \emph{For},
\emph{While}, \emph{Case} are abstractions in BOAST matching the behavior of
corresponding control structures in other languages. An exception is the For in
BOAST more closely matches the for in Fortran than the one in C.

Figure~\ref{fig:BOAST_control} shows some basic usages of the control structures.
The example shows a C macro or function that behaves similarly to the Fortran modulo
intrinsic. The sample script (inefficiently) computes and stores in \emph{j} the
number of multiples of $7$ in the 0 to 100 range. It uses the For and If control
structures. A Ruby block is passed to each of those constructs. This block is
evaluated at the time the construct is printed. If several such constructs are
needed (in an \emph{if elsif else} case, for instance) they can be explicitly
passed as parameters using the \emph{lambda} Ruby keyword.

@@

\lstset{style=customf,
        caption={Fortran output.},
        label={lst:BOAST_procedure_FORTRAN} }

@@

The last control structure is Procedure. It describes either procedures
or functions. Figure~\ref{fig:BOAST_procedure} presents the use of this
abstraction. It illustrates the signature of a real kernel from BigDFT. This kernel uses
an input array \emph{x} and an output array \emph{y} of double precision
numbers. Those arrays have two dimensions which depend on input variables
\emph{n} and \emph{ndat}. We can see here the use of the \emph{opn} and
\emph{close} keywords that are used to print a control structure without an
associated Ruby block. This time we placed ourselves inside of BOAST namespace.

The generated outputs in Fortran (Listing~\ref{lst:BOAST_procedure_FORTRAN}) and
C (Listing~\ref{lst:BOAST_procedure_C}) show the difference in meta-information that is
kept between both versions.

@@

\subsection{BOAST Run-time}

In the previous Section we presented BOAST's language. This allows us to
describe procedures and functions and to meta-program them using Ruby.
Each version has to be compiled, linked and executed to assess its performance
in order to find the best version of a computing kernel.  This can be very
time consuming if this process cannot be automated. By enabling more versions
for evaluation, the automation will bring improved portability, better performance,
and in the end will improve the productivity of the developer.

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects include:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

  \subsubsection{Multi-target Language Generation}
\label{sec:multitarget}
Language availability and performance varies between platforms. It is thus
important to express computing kernels in different languages, based
on the availability and their respective merits on the target platform. Some
languages have additional features, such as languages that target GPUs
(OpenCL, CUDA). The developer should be given tools to determine what kind of
language is currently used in order to be able to use those additional features.

@@

Compilation of the generated kernels must also be very flexible because HPC
application developers may encounter platforms with very diverse compilation environments.
Proprietary and dedicated compilers are common on HPC infrastructures. Thus
BOAST build system exhibits similar behavior to common build systems. Compilers
and their compile/build options can be specified at several
places. The list by increasing order of precedence includes: BOAST configuration
file, environment variables, and at kernel build time.

@@

The next logical step is to benchmark the built kernel. BOAST offers a simple
way to run a kernel that was successfully built while staying within BOAST.
A built BOAST \emph{CKernel} exposes a run method that accepts arguments corresponding
to the BOAST \emph{Procedure} used to instantiate the kernel. Arrays must be
instances of \emph{NArray} which are numerical arrays that use C arrays
underneath.

Arrays which correspond to output parameters will be modified during the
execution of the kernel so results can be checked. This allows for easy non
regression testing. The run method also returns information (and result for
kernels that are functions as well as output scalars) about the run. For
instance, one such information includes the run time of the kernel and it is
obtained using the system-wide real-time clock. Other probes can be inserted
at compile-time, if needed.
BOAST also supports PAPI~\cite{mucci1999papi} to capture hardware performance 
counters during each kernel execution.

@@

The non regression methodology presented before is viable provided input data
can be generated at run-time. For instance, in the case of the Laplace kernel,
generating an input image and the corresponding reference output image, using a
reference implementation, is easy. Unfortunately it is not always possible, some
applications have complex data patterns that are difficult to
synthesize without running the full application. Thus BOAST offers a way to load
binary data from the file system and use them as inputs of a kernel. Outputs can
also be checked against those binary data thus enabling (almost) data oblivious
non regression testing.

@@

Debugging applications running on GPU environments is well-recognized as a hard
and time-consuming activity. In complement with BOAST, we designed a trace-based
debugging tool that simplifies this porting operation. The tool relies on BOAST
support of multi-target code generation (Section~\ref{sec:multitarget}), used to
validate an application from one GPU-programming framework to another or between
different implementations using the same programming model. A case-study of the
port of a CUDA application to OpenCL is presented in
Section~\ref{subsec:specfem}.

@@

%
% HSG Question: why GPUTrace has to be state-full?
%
However, in contrast with \code{strace} and \code{ltrace},
\code{GPUTrace} has to be \emph{state-full}. Indeed, most of API
parameters are handles to opaque types, so, in order to provide
meaningful information, \code{GPUTrace} stores information about these
objects (handle value, buffer creation size and attributes, kernel
name and prototype, \etc{}). This is a custom trace format.

Once two call traces have been generated by \code{GPUTrace}, the user
can compare them with a graphical \code{diff} tool, and spot the
different porting mistakes: two parameters reversed, an offset
incorrectly applied, \etc{}

By default, \code{GPUTrace} only prints a unique identifier (the
creation index) for the memory buffers.  Additionally, \code{GPUTrace}
supports several modifier flags. One flag can be activated
to append the first bits of the buffer to the trace, printed in the
right format, for visual inspection. Another flag can be set to
dump the whole content of the buffer into a file, for a full
inspection.

@@

In this section we will present the benefits of using BOAST on the Laplace
motivating example as well as two scientific applications that use BOAST. The
first one, BigDFT~\cite{genovese2008daubechies}, uses BOAST in order to develop new
functionalities with performance portability in mind. The second one,
SPECFEM3D~\cite{komatitsch2011fluid}, uses BOAST to factorize OpenCL and CUDA
development, while having robust non regression tests.

@@

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimizations can have an impact on the
performance of the kernel on the Mali architecture. But, what is the impact in other
architectures? And, are there additional optimizations that can impact the
performance?

@@

Here are the parameters we finally selected for our kernel optimization and
their possible values:
\begin{itemize}
\item \textit{x\_component\_number}: a positive integer
\item \textit{y\_component\_number}: a positive integer
\item \textit{vector\_length}: 1, 2, 4, 8 or 16
\item \textit{temporary\_size}: 2 or 4
\item \textit{synthesize\_loads}: \textit{true} or \textit{false}
\item \textit{vector\_recompute}: \textit{true} or \textit{false}
\end{itemize}
The last parameter is used when \textit{x\_component\_number} is not divisible by
\textit{vector\_length}. Two solutions are possible then, divide the remainder
of the division in vectors of smaller sizes (\textit{vector\_recompute} =
\textit{false}) or load more data and compute useless values in vectors of the
specified size (\textit{vector\_recompute} = \textit{true}).
This last option mimics the behavior of the ARM implementation, although when
working at the component level it may not be a valuable thing to try.

@@

Table~\ref{tbl:ARM} show the best results obtained by ARM on different images
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically outperforms the hand
optimized version. As far as the optimization options are concerned, the results
are disappointing: the same kernel configuration is the best for all image
sizes. This kernel uses \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} because \textit{x\_component\_number}
== \textit{vector\_length}. Those results show that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in one cycle without
compromising the bandwidth. The results shown here are the best of four runs for
each configuration.

@@

But what if we run our benchmark on other architectures? Table~\ref{tbl:other}
shows the results obtained when running our BOAST implementation on other
architectures. The chosen architectures include an Intel i7-2760QM CPU (Sandy
Bridge architecture) that supports OpenCL 1.2 and a system with an NVIDIA gtx680 GPU that
supports OpenCL 1.1. We see that the performance ratio between the different
architectures is stable across image sizes.

The optimization parameters selected are not the same for those architectures.
Indeed, the Intel CPU favors kernels that have the parameters:
\textit{x\_component\_number} = 16,
\textit{vector\_length} = 8,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false. Once again
\textit{vector\_recompute} does not apply. \textit{y\_component\_number} varies
from 4 to 2 when image size increase, thus decreasing task granularity as the
global work size increase. This result is unexpected and understanding why this
happen could be interesting. The NVIDIA GPU favors processing square tiles:
\textit{x\_component\_number} = 4, \textit{y\_component\_number} = 4,
\textit{vector\_length} = 4, \textit{temporary\_size} = 2 and
\textit{synthesize\_loads} = false. Once more, \textit{vector\_recompute} has no 
meaning.
